\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2022-01-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

{\nib {\blue The original contribution of this article is to design an algorithm for \emph{online nonparametric} computation of the joint state PDFs subject to stochastic uncertainties (initial conditions, parameters, process noise and combinations thereof) in power system dynamics.

The qualifier ``online" means that it is not a variant of Monte Carlo where samples are first propagated \ul{and then} a function approximation is performed offline to estimate the joint state PDF at a snapshot. The qualifier ``nonparametric" means that neither the statistics nor the dynamics are approximated \emph{a priori}. This is novel because to the best of the author's knowledge, there does not exist a power systems paper which can accomplish the same. The state-of-the-art in power system literature is discussed in Sec I.1. More importantly, Sec. I.2 details why the state-of-the-art is what it is. \ul{We specifically explain the contribution of this article in Sec. I.3 titled ``Contributions of this paper".}
}}

\noindent
{\bf R1.2. Comments by reviewer}:\\
{\em The structure of the article should be highly improved. A large portion of the contexts is simply repeating the well-known power system classical model, kroon reduction, and the FPF equations.  Only from section IV, page 6, I can see some of the proposed weighted points methods.}

{\nib{ \blue We thank the reviewer for this comment. This paper is \emph{not} about novel implementation of an existing idea. The idea itself is new and consequently needs to be contextualized and developed in the paper. \ul{The discussion toward the end of Sec. II and that in Sec III are, in our opinion, more fundamental contribution than the implementation in Sec. IV. Those parts of the paper really mark the innovation since they explain why someone, even if conversant in infinite dimensional gradient flow literature, cannot directly apply existing techniques to this problem. Then Sec. III-C and III-D explain how to circumvent the specific difficulties arising in this problem. A reader jumping to Sec. IV assuming that everything before that is known, is mistaken}. Notice that we clearly laid out the structure of this paper (which section does what) in the paragraph right before Sec. II starts.}}

@@

\noindent
{\bf R1.3. Comments by reviewer}:\\
{\em I have major doubts about the model used in the article for the following reasons.\\
(1)     The paper claim the proposed method can propagate stochastic uncertainties in realistic power networks. However, only a classical model is used, which is not realistic at all.\\
(2)     Why do active power and reactive power from load have no randomness at all. This is too unrealistic.\\
(3)     The diffusion term of this article is only considered in the frequency. The process noise in the rotor angle is fully ignored. This is also unreasonable.\\
To improve your work, I highly suggested the authors follow [R1], which is missing in your literature but can help the authors better know Power system SDE.  Also, it is suggested to introduce the OU process to model the randomness in the load, which cannot be ignored in practice.\\
R1. A systematic method to model power systems as stochastic differential-algebraic equations, F. Milano, R. Zarate-Minano, IEEE Transactions on Power Systems, 28(4), 4537-4544.}

{\nib{ \blue Our responses to the three numbered sub-comments below.

(1) The second order sample path dynamics model we used is standard in power systems literature, and specifically as we pointed}}

@@

(ii) \ul{What we propose is not a model}. Transcribing the Fokker-Planck operator into a generalized gradient flow, and thereby deriving the corresponding proximal recursion, is a purely mathematical construct. There are no modeling assumptions made in that derivation. \ul{Our contribution in this paper is to initiate and showcase this new way of solving the transient Fokker-Planck PDE subject to a simple but popular power system dynamic model}. Extending this framework to higher order sample path dynamics is mentioned as future work, and is beyond the scope of this particular paper.

}}

\noindent
{\bf R3.2. Comments by reviewer}:\\
{\em The literature is incomplete, many recent works on SDAEs for power system applications are not mentioned.}

{\nib \blue{\ul{As per the reviewer's suggestion, in the revised manuscript, we have included two sentences (marked blue) in Sec. I.1, toward the end of second paragraph, mentioning the usage of SDAEs in power system dynamic simulation}. Therein, the first sentence contains two newly added references: [22] and [23], both on simulating SDAE sample paths.}}

### 2022-01-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf AE.1. Comments by AE}:\\
{\em The authors are requested to revise and resubmit their paper after addressing reviewer feedback. Of particular importance is that reviewers point out the model leveraged for analysis is overly simplistic.}

{\nib {\blue We are indebted to the AE and the reviewers for their help to improve the manuscript.

In addition to the itemized responses to the reviewers' comments and the corresponding revisions marked in blue in the manuscript, we have made the \ul{following additional fixes and improvements} (also marked in blue in the revised manuscript):\\
(i) In equation (8), we corrected a typo in its right hand side: the second summand runs over pairs $(i,j)$ such that $i<j$.\\
(ii)

{\red{Regarding the usage of simple model, please see our response to comment \textbf{R3.1} and \textbf{1.3, sub-comment (1)}. In summary, the current state-of-the-art in power system literature for this problem is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics.  \ul{We clarify two important points here}:\\(i) The proposed proximal recursion framework on the space of probability measures is not conceptually limited by the second order dynamics, and we expect follow up research to design custom proximal recursions for higher order synchronous machine models. In this sense, the present paper should be seen as the stepping stone laying the foundation for handling more complex dynamical models. The authors have been completely up front by explicitly mentioning this in the Conclusions section. We have also reiterated the same in this rebuttal.\\(ii) Deriving proximal recursion requires understanding the generalized gradient flow structure over probability measures. Our research disposition is that one cannot bypass understanding the generalized gradient flow structure for the simpler second order sample path dynamics before attempting the same for more complex sample path models. In fact, a significant contribution of our article, in the authors' opinion, is that it distills out what exactly is technically challenging and what are the peripheral issues.}} 
}}

@@

{\nib{ \blue We thank the reviewer for this comment. This paper is \emph{not} about novel implementation of an existing idea. The idea itself is new and consequently needs to be contextualized and developed in the paper. \ul{The discussion toward the end of Sec. II and that in Sec III are, in our opinion, more fundamental contribution than the implementation in Sec. IV. Those parts of the paper really mark the innovation since they explain why someone, even if conversant in infinite dimensional gradient flow literature, cannot directly apply existing techniques to this problem. Then Sec. III-C and III-D explain how to circumvent the specific difficulties arising in this problem. A reader jumping to Sec. IV assuming that everything before that is known, is mistaken}. Notice that we clearly laid out the structure of this paper (which section does what) in the paragraph right before Sec. II starts.}}

\noindent
{\bf R1.3. Comments by reviewer}:\\
{\em I have major doubts about the model used in the article for the following reasons.\\
(1)     The paper claim the proposed method can propagate stochastic uncertainties in realistic power networks. However, only a classical model is used, which is not realistic at all.\\
(2)     Why do active power and reactive power from load have no randomness at all. This is too unrealistic.\\
(3)     The diffusion term of this article is only considered in the frequency. The process noise in the rotor angle is fully ignored. This is also unreasonable.\\
To improve your work, I highly suggested the authors follow [R1], which is missing in your literature but can help the authors better know Power system SDE.  Also, it is suggested to introduce the OU process to model the randomness in the load, which cannot be ignored in practice.\\
R1. A systematic method to model power systems as stochastic differential-algebraic equations, F. Milano, R. Zarate-Minano, IEEE Transactions on Power Systems, 28(4), 4537-4544.}

{\nib{ \blue Our responses to the \ul{three numbered sub-comments} below.

(1) As pointed out in the last paragraph of Sec. II-B, the current state-of-the-art in power system literature for the problem we are addressing is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics.  \ul{We clarify two important points here}:\\ \textbf{[1a]} The proposed proximal recursion framework on the space of probability measures is not conceptually limited by the second order dynamics, and we expect follow up research to design custom proximal recursions for \ul{higher order synchronous machine models}. In this sense, the present paper should be seen as the stepping stone laying the foundation for handling more complex dynamical models. The authors have been completely up front by explicitly mentioning this in the Conclusions section. \\ \textbf{[1b]} Deriving proximal recursion requires understanding the generalized gradient flow structure over probability measures. \ul{Our research disposition is that one cannot bypass understanding the generalized gradient flow structure for the simpler second order sample path dynamics before attempting the same for more complex sample path models}. In fact, a significant contribution of our article, in the authors' opinion, is that it distills out what exactly is technically challenging and what are the peripheral issues.

(2) The }}

### 2022-01-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

{\red{Regarding the usage of simple model, please see our response to comment \textbf{R3.1} and \textbf{1.3, sub-comment (1)}. In summary, the current state-of-the-art in power system literature for this problem is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics.  \ul{We clarify two important points here}:\\(i) The proposed proximal recursion framework on the space of probability measures is not conceptually limited by the second order dynamics, and we expect follow up research to design custom proximal recursions for higher order synchronous machine models. In this sense, the present paper should be seen as the stepping stone laying the foundation for handling more complex dynamical models. The authors have been completely up front by explicitly mentioning this in the Conclusions section. We have also reiterated the same in this rebuttal.\\(ii) Deriving proximal recursion requires understanding the generalized gradient flow structure over probability measures. Our research disposition is that one cannot bypass understanding the generalized gradient flow structure for the simpler second order sample path dynamics before attempting the same for more complex sample path models. In fact, a significant contribution of our article, in the authors' opinion, is that it distills out what exactly is technically challenging and what are the peripheral issues.}} 
}}

\noindent
{\bf AE.2. Comments by AE}:\\
{\em Authors are also encouraged to revisit their overview of prior art.}

{\nib {\blue Following the suggestions, the revised manuscript has expanded the overview of prior art. Specifically, we have included \ul{two new references [22], [23] on stochastic differential algebraic equations (SDAEs) and cited them inline in Sec. I.1, second paragraph, as detailed in our response to the comment \textbf{R3.2} in this document}. This includes the reference suggested in comment \textbf{R1.3}. \ul{We have also included the reference suggested in comment \textbf{R4.4}; this is reference [28] in the revised manuscript. Its inline citation is in Sec. I.1, last paragraph, first sentence}.}}

@@

{\nib{ \blue We thank the reviewer for this comment. This paper is \emph{not} about novel implementation of an existing idea. The idea itself is new and consequently needs to be contextualized and developed in the paper. \ul{The discussion toward the end of Sec. II and that in Sec III are, in our opinion, more fundamental contribution than the implementation in Sec. IV. Those parts of the paper really mark the innovation since they explain why someone, even if conversant in infinite dimensional gradient flow literature, cannot directly apply existing techniques to this problem. Then Sec. III-C and III-D explain how to circumvent the specific difficulties arising in this problem. A reader jumping to Sec. IV assuming that everything before that is known, is mistaken}. Notice that we clearly laid out the structure of this paper (which section does what) in the paragraph right before Sec. II starts.}}

\noindent
{\bf R1.3. Comments by reviewer}:\\
{\em I have major doubts about the model used in the article for the following reasons.\\
(1)     The paper claim the proposed method can propagate stochastic uncertainties in realistic power networks. However, only a classical model is used, which is not realistic at all.\\
(2)     Why do active power and reactive power from load have no randomness at all. This is too unrealistic.\\
(3)     The diffusion term of this article is only considered in the frequency. The process noise in the rotor angle is fully ignored. This is also unreasonable.\\
To improve your work, I highly suggested the authors follow [R1], which is missing in your literature but can help the authors better know Power system SDE.  Also, it is suggested to introduce the OU process to model the randomness in the load, which cannot be ignored in practice.\\
R1. A systematic method to model power systems as stochastic differential-algebraic equations, F. Milano, R. Zarate-Minano, IEEE Transactions on Power Systems, 28(4), 4537-4544.}

{\nib{ \blue \ul{Our responses to the sub-comments below}.

(1) \textbf{Regarding usage of classical model:} Whether a model is relevant depends on the problem context. As pointed out in the last paragraph of Sec. II-B, the current state-of-the-art in power system literature for the problem we are addressing is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics.  \ul{We clarify two important points here}:\\ \textbf{[1a]} The proposed proximal recursion framework on the space of probability measures is not conceptually limited by the second order dynamics, and we expect follow up research to design custom proximal recursions for \ul{higher order synchronous machine models}. In this sense, the present paper should be seen as the stepping stone laying the foundation for handling more complex dynamical models. \ul{The authors have been completely up front by explicitly mentioning this in the Conclusions section}. \\ \textbf{[1b]} Deriving proximal recursion requires understanding the generalized gradient flow structure over probability measures. \ul{Our research disposition is that one cannot bypass understanding the generalized gradient flow structure for the simpler second order sample path dynamics before attempting the same for more complex sample path models}. In fact, a significant contribution of our article, in the authors' opinion, is that it distills out what exactly is technically challenging and what are the peripheral issues.

(2) \textbf{Regarding randomness in active and reactive power:} {\red{TBD}}

(3) \textbf{Regarding the diffusion term:} The reason why the diffusion term is present in the RHS of (1b) but not in the RHS of (1a) is because (1) is the state space representation of the second order Langevin dynamics (6). For the Langevin dynamics (6), its corresponding It\^{o} SDE \ul{has to be} (1). This is simply re-writing of a second order differential equation as an equivalent system of first order differential equations. There is no extra assumption here. \ul{Indeed, it is mathematically incorrect to add a noise term in the RHS of (1a) while claiming consistency with (6)}.

\ul{Notice that the absence of additive noise in the RHS of (1a) does \emph{not} imply that the rotor angles are deterministic trajectories}. The reason why the rotor angles are stochastic processes is because the process noise in (1b) affects the rotor angle sample paths via dynamical nonlinear coupling with the angular velocities, i.e., these are second order stochastic effects. \ul{In short, the noise in rotor angles is \emph{not} ignored; in fact the rotor angles are nonlinearly colored (i.e., non-white) noisy sample paths}.

This reviewer may also have \ul{mathematical misconception} in thinking that a model that has additive noise in the RHS of \emph{all} SDE components is somehow more general than the case where the additive noise terms are absent from some but not all components of an SDE RHS. \ul{In reality, the complexity is in the other way around}. The SDEs of the form (1) are the so-called ``degenerate It\^{o} SDEs" which are much more difficult to analyze than the non-degenerate versions (noise in the RHS of all SDE components). For instance, this degeneracy makes the joint PDF evolution equation (10) a \emph{kinetic} Fokker-Planck PDE, which is different in structure than the standard Fokker-Planck a.k.a. Kolmogorov's forward PDE. \ul{The gradient flow structure and the associated measure-valued proximal recursions for the kinetic Fokker-Planck PDE are less understood than the same for the standard Fokker-Planck PDE. This is exactly why the methodological contribution of this paper is important}.

\textbf{Regarding the reference [R1]:} \ul{In the revised manuscript, we have included the ref. [R1] suggested by the reviewer with inline citation in Sec. I.1, toward the end of second paragraph. Please note that the topic of [R1] is different from the topic of our paper: how to perform online nonparametric computation of transient joint state PDFs}.

\textbf{Regarding OU process in load:} \ul{It appears that the reviewer have missed what is actually going on in our paper}. The Kron reduced second order stochastic model implies that in the unreduced sense, the load dynamics was first order noisy Kuramoto, in addition to the synchronous machine dynamics being second order noisy Kuramoto. Therefore, we do have stochastic load dynamics, which is in fact more realistic than OU process model. While OU process assumes linear drift with constant diffusion, our setting considers the load SDEs having drifts with trigonometric nonlinearity and constant diffusion.

### 2022-01-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

{\nib {\blue The original contribution of this article is to design an algorithm for \emph{online nonparametric} computation of the joint state PDFs subject to stochastic uncertainties (initial conditions, parameters, process noise and combinations thereof) in power system dynamics.

The qualifier ``online" means that it is not a variant of Monte Carlo where samples are first propagated \ul{and then} a function approximation is performed offline to estimate the joint state PDF at a snapshot. The qualifier ``nonparametric" means that neither the statistics nor the dynamics are approximated \emph{a priori}. This is novel because to the best of the author's knowledge, there does not exist a power systems paper which can accomplish the same. The state-of-the-art in power system literature is discussed in Sec I.1. More importantly, Sec. I.2 details why the state-of-the-art is what it is. \ul{We specifically explain the contribution of this article in Sec. I.3 titled ``Contributions of this paper".}
}}

\noindent
{\bf R1.2. Comments by reviewer}:\\
{\em The structure of the article should be highly improved. A large portion of the contexts is simply repeating the well-known power system classical model, kroon reduction, and the FPF equations.  Only from section IV, page 6, I can see some of the proposed weighted points methods.}

{\nib{ \blue We thank the reviewer for this comment. This paper is \emph{not} about novel implementation of an existing idea. The idea itself is new and consequently needs to be contextualized and developed in the paper. \ul{The discussion toward the end of Sec. II and that in Sec III are, in our opinion, more fundamental contribution than the implementation in Sec. IV. Those parts of the paper really mark the innovation since they explain why someone, even if conversant in infinite dimensional gradient flow literature, cannot directly apply existing techniques to this problem. Then Sec. III-C and III-D explain how to circumvent the specific difficulties arising in this problem. A reader jumping to Sec. IV assuming that everything before that is known, is completely mistaken}. Notice that we clearly laid out the structure of this paper (which section does what) in the paragraph right before Sec. II starts.}}

@@

\ul{Notice that the absence of additive noise in the RHS of (1a) does \emph{not} imply that the rotor angles are deterministic trajectories}. The reason why the rotor angles are stochastic processes is because the process noise in (1b) affects the rotor angle sample paths via dynamical nonlinear coupling with the angular velocities, i.e., these are second order stochastic effects. \ul{In short, the noise in rotor angles is \emph{not} ignored; in fact the rotor angles are nonlinearly colored (i.e., non-white) noisy sample paths}.

This reviewer may also have \ul{mathematical misconception} in thinking that a model that has additive noise in the RHS of \emph{all} SDE components is somehow more general than the case where the additive noise terms are absent from some but not all components of an SDE RHS. \ul{In reality, the complexity is in the other way around}. The SDEs of the form (1) are the so-called ``degenerate It\^{o} SDEs" which are much more difficult to analyze than the non-degenerate versions (noise in the RHS of all SDE components). For instance, this degeneracy makes the joint PDF evolution equation (10) a \emph{kinetic} Fokker-Planck PDE, which is different in structure than the standard Fokker-Planck a.k.a. Kolmogorov's forward PDE. \ul{The gradient flow structure and the associated measure-valued proximal recursions for the kinetic Fokker-Planck PDE are less understood than the same for the standard Fokker-Planck PDE. This is exactly why the methodological contribution of this paper is important}.

\textbf{Regarding the reference [R1]:} \ul{In the revised manuscript, we have included the ref. [R1] suggested by the reviewer with inline citation in Sec. I.1, toward the end of second paragraph. Please note that the topic of [R1] is different from the topic of our paper. We are interested in how to perform online nonparametric computation of transient joint state PDFs}. This is not Monte Carlo.

\textbf{Regarding OU process in load:} \ul{It appears that the reviewer might have missed what is actually going on in our paper}. The Kron reduced second order stochastic model implies that in the unreduced sense, the load dynamics was first order noisy Kuramoto, in addition to the synchronous machine dynamics being second order noisy Kuramoto. Therefore, we do have stochastic load dynamics, which is in fact more realistic than OU process model. While OU process assumes linear drift with constant diffusion, our setting considers the load SDEs having drifts with trigonometric nonlinearity and constant diffusion.

### 2022-01-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf R1.6. Comments by reviewer}:\\
{\em The simulation time is too short to convince the readers. Only 1 second is considered here. Note that the errors can be quickly accumulated in the iteration.}

{\nib{ \blue TBD.}}

\noindent
{\bf R1.7. Comments by reviewer}:\\
{\em It is suggested to count the total computing time, not on each step.}

{\nib{ \blue While the depiction of the computational times in Figs. 8 and 9 in the revised manuscript are somewhat unconventional, we explain why we choose to plot that way. \ul{To aid clarity, in the revised manuscript, we have added an excerpt of the following at the end of Sec. V-A.}

@@

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{in our opinion, this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

\emph{Second}, \ul{the planar contour marginal plots require additional mental efforts from the readers to correctly parse the statistical information conveyed}. For instance, consider the fourth column, second row subplot in Fig. 1 in this rebuttal document. It is upon the reader to parse that the two high probability regions on the left and right of that subplot are actually referring to the same angular high probability region around $\theta_{2}=0$ rad, thanks to the periodic boundary condition on the angular variable. Our philosophy of technical communication is that the onus is on the authors, not on the readers, to clarify the scientific information as correctly as possible. And yes, we are aware that many papers in the power system literature make plots in the rectangular domain ignoring the manifold nature of the support, which in our experience, continue to add pain to the graduate students and beginning researchers trying to decipher the statistical meaning conveyed in those plots. We thus prefer to make plots that depict the variables as they are.}}
